% !TeX program = pdflatex

\documentclass{utthesis}

\usepackage{lipsum} % Dummy text for examples
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{longtable}
\usepackage{amsmath, amssymb}
\usepackage{url}
\usepackage{subfigure}
\usepackage{afterpage}
\usepackage{pdflscape} % for 'landscape' environment
\usepackage[pagebackref]{hyperref} % Handy for links, can comment for plain PDF
\bibliographystyle{aasjournal}

\DeclareMathOperator\erf{erf}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}

\UTyear=2016

\makeindex

%\graphicspath{{./Figures/}}


\begin{document}

\author{Kevin Gullikson}
\title{Spectroscopic Detection and Characterization of Extreme Flux-Ratio Binary Systems}
\date{Revised: \today}

\UTcopyrightlegend % Optional

\begin{UTcommittee}
\UTaddsupervisor{Adam Kraus}
\UTaddcommittee{Sarah Dodson-Robinson}
\UTaddcommittee{Daniel T. Jaffe}
\UTaddcommittee{Edward L. Robinson}
\UTaddcommittee{Michael Meyer}
\end{UTcommittee}

\UTtitlepage{M.A.}{May}

\frontmatter


\begin{UTdedication}
Dedicated to my loving parents, who encouraged my exploration of science from an early age.
\end{UTdedication}

\cleardoublepage
\setcounter{page}{5}

\begin{UTacknowledgements}
I would like to start by thanking everyone who served as a supervisor during part of my thesis. Mike Endl served as my co-supervisor for two years, taught me how to observe on the 2.7m telescope at McDonald observatory, and involved me in the planet search program for the last 5 years. That experience provided many interesting nights, a fun "side project", and gave me the confidence to do much of the observing for my main survey program. Sally Dodson-Robinson served as my thesis supervisor for my first two years. She gave me a great deal of independence in my project, while providing valuable feedback. I would like to thank both her and my current supervisor, Adam Kraus, for nominating me for several UT fellowships and funding me as a research assistant during much of my graduate career. The freedom to work exclusively on research has been invaluable, and facilitated much of the progress that I made. Adam, thank you for being a great advisor. Despite being brand new at this, you provide a great mix of a hands-on style when I need it and hands-off when I don't, and have been nothing but supportive in every way. 

I would also like to thank Rob Robinson, my committee member, for numerous discussions on statistical methods both while planning and analyzing the results of my survey program. Discussions with you, not to mention your excellent data analysis class, have really helped me to think of statistical methods in a way that makes sense.

Finally, I would like to thank all of my friends and colleagues at UT for making these last years so much fun. From the usual crowd at Crown to cookouts at Paul's to boat parties and volleyball with Tom Montemayor, I have rarely lacked for a way to blow off some steam and relax in between the stressful moments. And of course, I would like to thank my girlfriend Cori Norman. You have made the last three years absolutely wonderful.

\end{UTacknowledgements}

%\include{acknowledgements}
%\include{preface}

\begin{UTabstract}{Adam Kraus}
Binary stars and higher-order multiple systems are a ubiquitous outcome of star formation, especially as the system mass increases. The companion mass-ratio distribution is a unique probe into the conditions of the collapsing cloud core and circumstellar disk(s) of the binary fragments. Inside $a \sim 1000$ AU the disks from the two forming stars can interact, and additionally companions can form directly through disk fragmentation. We should therefore expect the mass-ratio distribution of close companions to differ from that of wide companions. This prediction is difficult to test with intermediate-mass primary stars using traditional methods because the contrast ratios that would be required to detect low-mass companions at narrow working angles are not yet achievable. In this thesis, we present a spectroscopic method to detect and characterize close companions to a variety of stars. We demonstrate applications of the method to detection of stars and even planets around sun-like stars, and present the results of a survey searching for companions to A- and B-type stars. As part of the survey, we estimate the temperatures and surface gravity of most of the 341 sample stars, and derive their masses and ages. We additionally estimate the temperatures and masses of the 64 companions we find, 23 of which are new detections. We find that the mass-ratio distribution for our sample has a very different form than in previous work, where it is usually well-described by a power law, and indicates that close companions to intermediate-mass stars experience significantly different accretion histories or formation mechanisms than wide companions.
\end{UTabstract}


% add . to include section numbers (non-standard)
\tableofcontents
%\tableofcontents.

\listoffigures

\listoftables

\mainmatter

\chapter{Introduction}

Stellar multiplicity is an inevitable and common outcome of star formation, with roughly half of all solar-type field stars in binary or multiple systems \citep{Raghavan2010} and an even higher fraction as the stellar mass increases \citep{Zinnecker2007}. Young stellar associations and clusters tend to have even higher multiplicity \citep{Duchene2013}, indicating that stars often form in multiple systems that are subsequently destroyed by dynamical interactions as the cluster dissociates. 

The overall multiplicity rate and the distributions of mass ratio, period, and eccentricity of a binary star population place important constraints on the mode of binary star formation. While the period and eccentricity are altered by dynamical processing in the birth cluster, the present-day mass ratio of a binary system is a direct result of its formation \citep{Parker2013}. Most binary stars are thought to form via core fragmentation \citep{Boss1979, Boss1986, Bate1995}, in which a collapsing core fragments into two or more individual protostars. The number and initial masses of the fragments are set by the total core mass, as well as its rotation, turbulence, and its temperature and density structure. \citet{Machida2008} performed a parameter study by simulating several collapsing clouds with different levels of thermal, rotational, and magnetic energies. They found that the initial fragments occur on wide separations ($a \gtrsim 3 AU$) if the rotational energy dominates that of the magnetic field. If the magnetic energy dominates, however, then the fragmentation occurs much later and generates very close binary systems ($a \lesssim 1 AU$). 

\citet{Machida2008} do not follow the accretion or migration evolution of the fragments, but we can estimate that evolution from the relevant physics: If the fragments are well separated ($a \gtrsim 1000$ AU), they will evolve independently of each other, accreting mass from the core material onto their own protostellar disks and then onto the protostars themselves.  However, close fragments ($a \sim 100$ AU) will interact with each other; the protostellar disk may be truncated, destabilized, or form into a circumbinary disk if the separation is small enough \citep{Bate1997}. In addition, an unstable disk can fragment to form low-mass close companions \citep{Kratter2006, Stamatellos2011} directly. The mass ratios of close companions formed via either mechanism should be affected by preferential accretion. Most work has suggested that the disk material and high angular momentum infalling core material will preferentially accrete onto the lower mass companion as it migrates inward \citep{Bate1997, BBB2002}, although some recent work has suggested the preferential accretion may go in the opposite direction as a result of magnetic disk braking \citep{Zhao2013}.

Beyond informing star formation models, intermediate-mass stars have recently seen a revival of interest as potential young planet hosts, spurred largely by the detection of planets orbiting nearby \mbox{$\sim2 M_{\odot}$} stars on both wide \citep[e.g.][]{Lagrange2010, Marois2008} and close \citep{Johnson2011} orbits. Since the main sequence lifetime of an A- or B-type star is tens to hundreds of Myrs, a planetary companion would still be bright and easier to detect with direct imaging techniques than the same companion orbiting an old, solar-type star \citep{Marley2007}. Robust age estimates for nearby intermediate-mass stars are therefore very important to provide a vetted sample of young systems to search, and to convert the companion brightness into a mass estimate. 

In the context of planetary companions, stellar-mass binary companions are contaminants; companions complicate radial velocity planet searches because they necessitate simultaneous modeling of both stellar motions \citep[e.g.][]{Bergmann2015}. Likewise, companions complicate direct imaging planet searches by requiring either extremely high-contrast instrumentation \citep{Thalmann2014} or specialized coronagraphs \citep{Crepp2010}. 

However, known binary stars are typically avoided in planet search programs for a more fundamental reason: the binary companion depletes or destroys the planet-forming disk. By combining a binary census of the $\sim 2$ Myr Taurus-Auriga star-forming region with a disk census of the same, \cite{Kraus2012} showed that close ($\lesssim 40$ AU) binaries are about 2-3 times less likely to host a protoplanetary disk, and so hasten disk dispersal. Even if a disk survives, it tends to be depleted in mass by a factor of $\sim 25$ for binary separations $\lesssim 30$ AU \citep{Harris2012}. A full binary census focusing on companions within $\sim 100$ AU is therefore necessary in order to generate a direct-imaging planet search sample.

\section{Detection Methods}

There are three methods traditionally used to search for companions to stars, whether they be stellar or planetary companions. The first is direct imaging, in which we take an image of the star and look for nearby point sources in the image. This can be either seeing-limited, in which case it is difficult to find companions inside $1''$ of the primary, or with adaptive optics systems on large-aperture telescopes, in which case the inner working angle is near the diffraction limit of the telescope. We can detect companions out to a few hundred mas from the primary with adaptive optics imaging methods \citep[see][for typical sensitivity curves of a binary survey]{DeRosa2014}, and even closer with more time-consuming and complicated observational techniques such as angular differential imaging \citep{Marois2006}. 

The second detection method is interferometry, in which multiple beams of light from the source are allowed to constructively or destructively interfere with each other before recording the light. In some cases, interferometry is done with multiple telescopes separated by several tens of meters, producing an effective aperture much larger than can be produced as a single mirror. In other cases, a mask is introduced in front of a single telescope to only allow light from certain beams through \citep[aperture masking,][]{Tuthill2000, Ireland2008}. Interferometry can usually achieve smaller working angles than imaging, but cannot achieve as high contrast \citep[see e.g.][]{Aldoretta2015}.

The final traditional method useful in searching for stellar companions is radial velocity monitoring, in which the radial velocity of the star is measured many times over periods of years to decades. An unseen companion will cause the radial velocity of the primary star to oscillate. Unlike the previous two methods, the radial velocity method has no inner working angle; in fact, it works best at finding very close companions. The sensitivity falls off as the companion orbital separation increases, because the signal induced on the primary star decreases. The sensitivity also falls off as the number of spectral lines in the primary star spectrum falls, and as it rotation speed ($v\sin{i}$) increases. Careful radial velocity monitoring surveys are capable of detecting motion as small as $\sim 1 \mathrm{m\ s}^{-1}$ when observing slowly rotating solar-type stars \citep[e.g.][]{Wittenmyer2006, Fischer2009, Pepe2011}, but have difficulty achieving better than $\sim 1 \mathrm{km\ s}^{-1}$ when observing rapidly rotating hot stars \citep{Becker2015}.

Both imaging and interferometry can usually characterize the companion in a single image, provided the primary star parameters are sufficiently well-known. However, radial velocity monitoring is typically only capable of measuring the mass function of the binary system:
\begin{equation}
 f(M_1, M_2) = \frac{(M_2\sin{i})^3}{(M_1+M_2)^3}
 \end{equation}
 and only after monitoring the system for $1-2$ orbital periods. For companions on long-period ($\sim 10$ years) orbits, it is an extremely time-consuming method. In Chapter \ref{chap:dsd}, we introduce a spectroscopic method that is capable of both detecting and characterizing close companions in single observations.


\section{Previous Observational Results}

The mass ratio, period, and eccentricity distributions are fairly well-known for solar type stars \citep{Duquennoy1991, Raghavan2010} and cooler stars \citep{Fischer1992, Delfosse2004}. Interestingly, the mass-ratio distribution appears to be invariant to separation for these stars \citep{Meyer2013}, contrary to the theoretical expectations. All of the distributions are much less certain for more massive stars. The reason for this is two-fold: first, more massive stars tend to be more rare and farther away, meaning many of the companions are angularly close to the very bright primaries and difficult to detect with imaging techniques. Second, the primary stars tend to be rapid rotators, which limits radial velocity precision to $\sim 1 \mathrm{km\ s}^{-1}$. Additionally, detecting the companion spectrum suffers from the same flux-ratio difficulties as imaging methods, and is the only way to characterize the detected companion.

Nonetheless, \citet{DeRosa2014} performed an adaptive optics imaging survey of nearby A-type stars, and found that the mass-ratio distribution is well-described by a power law with large slope, indicating a very strong preference for low-mass companions. They also found initial evidence that the mass-ratio distribution for companions inside $125$ AU has a much shallower power law slope than that of wide companions, and is consistent with flat. Their close companion subsample contained only 18 binary systems, and the result is complicated by the inherent difficulty of detecting close companions with low mass ratios in an imaging survey. 

Radial velocity monitoring surveys can detect much closer companions than imaging surveys, but are typically only complete to low-mass companions if the primary is a slow rotator. Chemically peculiar Am stars are typically slow rotators, a fact typically attributed to tidal braking from binary companions; they thus form a highly biased sample of intermediate-mass stars. Nonetheless, it is interesting to note that they have a mass-ratio distribution which peaks near $q \sim 0.5$ \citep{Vuissoz2004}, an entirely different form than the power law distribution found around chemically normal stars at wide separations.

TODO: Discuss how the thesis is organized!


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Chapter 1  (DSD method)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\include{paper5}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Chapter 2 (telfit)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\include{paper3}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Chapter 3 (Ps1 Dra A search)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\include{paper4}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Chapter 4 (IGRINS simulation)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\include{paper2}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Chapter 5  (B-star pilot study)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\include{paper1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Chapter 6  (main survey)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\include{paper6}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Chapter 7  (Conclusions)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Conclusions}

In this thesis, we developed and characterized a spectral method to find companions to stars with a variety of temperatures. The method can be summarized as a two step process: first, prepare the spectrum by attempting to remove the signal from the primary star spectrum. The second step is to cross-correlate the prepared spectra against a grid of model spectra spanning a large parameter space. The cross correlation function (CCF) attempts to match the model spectrum template against the prepared data; if it finds a match for the template in the data then the CCF will have a strong peak at the radial velocity of the companion.

We present two different ways to prepare the spectrum. In Chapter \ref{chap:dsd}, we discuss the method we use for our main survey program (Chapters \ref{chap:pilot} and \ref{chap:survey}), which takes advantage of the fact that the primary star spectrum is very simple compared to the expected companion spectrum. In this case, we prepare the spectra for a companion search by simply subtracting a low-pass filtered version of the original spectrum, a process known as unsharp masking. We extensively characterize the direct spectral detection method for this type of primary star, estimating the companion sensitivity as well as the accuracy of the companion parameters we derive. We find that our method is capable of detecting late-K to early-M spectral type companions to A0V primary stars, and can detect them \emph{independent of the orbital separation}. As the primary star increases in mass and temperature, the coolest detectable companion gets hotter; however, the lowest detectable mass ratio decreases. 

We search for a bias in the derived parameters of the companion star by simulating binary star observations with real data, and find that the estimated temperature tends to be larger than the true temperature. The bias is usually largest at the coolest temperatures where the model spectra are least accurate due to the difficulty of modeling complex molecular absorption features. We calibrate the bias for four different spectrographs using a linear fit, which we use to convert from measured to actual temperature.

The second way for preparing the spectrum is by using orbital information and a large degree of orbital phase coverage to remove the signal from the primary star spectrum. We describe this method in Chapters \ref{chap:psi1dra} and \ref{chap:hotjup}. The basic idea is to take advantage of the fact that the spectral lines from the primary and secondary move in different directions. Therefore if you align several spectra from different epochs such that the primary star lines are in a constant position, the companion spectral lines will be washed out. The result is an empirical estimate of the primary star spectrum that is only minimally contaminated by the companion spectrum. We remove the primary star contamination from each individual spectrum by subtracting the empirical estimate, leaving only a spectrum of the companion star. Unlike the method we describe above and in Chapter \ref{chap:dsd}, this method requires several observations of the system that span a large fraction of the orbit; it is therefore not independent of separation. The great advantage of this method is that it can detect cool companions even when the primary star spectrum is complicated, and can detect much more extreme flux-ratio companions. In Chapter \ref{chap:hotjup}, we show that the method is even capable of detecting planet-mass companions, which entails a contrast ratio of $\sim 10^4$ in the K-band.

\section{The Separation-Variant Companion Mass-Ratio Distribution}

The bulk of the work in this thesis was motivated by the following question: is the companion mass-ratio distribution for binary systems with close ($a \lesssim 100 AU$) orbits significantly different from companions on wide ($a \gtrsim 100 AU$) orbits? Since we care about a quantity that changes with separation, we want a method that detects companions in such a way that the completeness does not depend on separation. Additionally, we want a method that can detect and characterize the binary system quickly. The direct spectral detection method described in Chapter \ref{chap:dsd} is just such a method, and we apply it to a binary survey in Chapter \ref{chap:survey} after performing a pilot survey with archived data in Chapter \ref{chap:pilot}.

We find that the companions we detect, which are necessarily all close companions because the light from both components must fall on the spectrograph slit, show a preference for companions with $q \sim 0.3$. The mass-ratio distribution for wide companions is well-fit by a power law with slope $\sim -2.3$ \citep{DeRosa2014}, indicating a very strong preference for low-mass companions. In contrast, our result is poorly fit by a power law, and is much better fit by a lognormal distribution. Our result is consistent with the inner mass-ratio distribution found by the VAST survey \citep{DeRosa2014}, an adaptive optics imaging survey. However, our distribution contains far more companions since it is specifically optimized for finding close, low-mass companions. We are therefore able to determine that not only does the slope of the mass-ratio distribution change with separation, the \emph{shape} of the distribution changes.

\section{Theoretical Implications}

The mass-ratio distribution derived in this work has a very different form than the power law found for companions at wide separations. This is likely a result of disk interactions as the two components are accreting. The close companions that we detect may form with similar masses to their counterparts at large separations ($a \gtrsim 1000\ AU$), but preferentially accrete matter from the dense primary star disk. The result would be a depletion of low mass ratio companions as they become intermediate to high mass ratio companions. The characteristic scale of $\sim 0.3$ that we see in Figure \ref{paper6_fig:mrd} would then be related to the disk timescale and the time at which the fragments form, since with enough time the preferential accretion would push all companions to $q = 1.0$. 

It is also possible that some of the companions found in this work were formed from a gravitationally unstable disk \citep[e.g.][]{Kratter2006, Stamatellos2011}. Being a completely different formation mechanism than the way wide companions form, we would expect the initial mass function to differ. Such companions would undergo the same preferential accretion discussed above. This explanation may even be preferable, since it could explain why we see a separation-dependent mass-ratio distribution around intermediate-mass stars but not low-mass stars: more massive stars tend to have more massive disks \citep{Andrews2013} that are more likely to fragment \citep{Kratter2010}.



\section{Future Work Needed}

Large scale simulations are likely needed to distinguish between the two formation scenarios and fully interpret the results of this survey. A significant amount of work has already been put towards this end in the form of radiation hydrodynamic simulations of giant molecular clouds \citep{Bate2012, Krumholz2012}. However, the present simulations do not generate enough stars more massive than the sun to quantitatively compare binary and multiple star statistics to observations. Additionally, there is a strong need for high resolution simulations to follow the full accretion and migration history of both core and disk fragments in order to compare the final mass ratios generated to observational studies such as the one presented here.

There is also a need for further observational efforts. Many of the companions we present in Chapter \ref{chap:survey} are unconfirmed, and so need at minimum follow-up spectroscopy. Follow-up imaging of more of the detections with an adaptive optics system would be useful as well, both for confirmation and validation of the companion temperature and to provide an estimate of the orbital separation. Alternatively or in addition to the imaging, the detections could be observed with spectroscopic measurements over several years to map out the full double-lined spectroscopic orbit and provide a dynamical mass ratio and separation measurement. 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%\begin{appendix}

%\chapter{First}
%\lipsum[3]

%\chapter{Second}

%\end{appendix}

\backmatter

\printindex

\cleardoublepage
\phantomsection
%\addcontentsline{toc}{chapter}{Bibliography}
%\chapter*{Bibliography}
%\lipsum[2]
\bibliography{references}

\end{document}
